---
title: "SOCI 245 - Week 10 Workshop"
author: "Matt Lawrence"
date: "4/27/20"
output: pdf_document
---

## Getting Started

To start, let's load the R packages we will need. Recall that in an R Markdown Notebook, you can click the green arrow in the chunk below to execute the commands.

```{r load packages and data, warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)

colleges <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci245/s20/data/colleges.csv")

colleges <- filter(colleges, TRUE)
colleges$name <- as.character(colleges$name)
colleges$public <- as.factor(colleges$public)
```


## Descriptive Statistics

We'll start by seeing how the *New York Times* uses Chetty et al's big dataset for their analyses. One thing to note: the *NYT* only used the 1981 cohort from Chetty's dataset; we'll use data pooled from the 1980-1982 cohorts. Most of the overall patterns we will see are similar to what the *NYT* reported but a few numbers will be slightly different. We will also restrict our dataset to four-year colleges.

We want to find the proportion of students at every college whose parents are in the top 1% of the income distribution and the proportion of students at every college whose parents are in the bottom 60% of the income distribution.

There is a variable for the proportion of students from the top 1%: `par_top1pc`. Let's find that value for Middlebury College. Recall that we can use indexing [with brackets] to find values of variables for observations that meet certain conditions. In this case, we want to pull the value of the `par_top1pc` variable from the `colleges` data frame for the observation where the value of the `name` variable is equal to "Middlebury College":

```{r top 1%}
colleges$par_top1pc[colleges$name=="Middlebury College"]
```

So 21% of Middlebury College's students in this sample come from the top 1% of the distribution of parents' income. The *NYT* chose to focus on the top 1%. The dataset also includes variables for the proportion of students from the top 5% (`par_top5pc`) and the top 10% (`par_top10pc`). Try to find the proportion of students at Middlebury from the top 5% and the proportion of students at a different institution from the top 10%. To find a different institution, the name and the case (upper, lower) have to be identical to how they are saved in the data frame. It is a good idea to view the data frame (use the spreadsheet icon to the right of the dataframe details in the upper right pane's environment tab), and browse the "Name" column. In the spreadsheet view, you can search for a college using the search bar (look for the magnifying glass in the top right). You can also sort by state, region, or selectivity tier.

### REPLACE THIS LINE WITH YOUR CODE ###


After the top 10% of the income distribution, students are grouped by income quintile from quintile 1 (the bottom 20%) to quintile 5 (the top 20%). The variable for the proportion of students from the bottom 20% is `par_q1` and the variable for the proportion of students from the top 20% is `par_q5` with the other three quintiles in the middle. How would we use these variables to create a new variable called `par_bottom60pc` which is the proportion of students from the bottom 60%?

```{r bottom 60%}
colleges$par_bottom60pc <- colleges$par_q1 + colleges$par_q2 + colleges$par_q3
```

What proportion of Middlebury College students come from families in the bottom 60% of the income distribution?

### REPLACE THIS LINE WITH YOUR CODE ###


Now let's calculate the statistic that the *NYT* analysis presents as its main finding, the ratio of students from the top 1% to the bottom 60%:

```{r calculating ratio}
colleges$ratio <- colleges$par_top1pc / colleges$par_bottom60pc
```

What is this value for Middlebury?

### REPLACE THIS LINE WITH YOUR CODE ###

```{r ratio for Middlebury}
colleges$ratio[colleges$name=="Middlebury College"]
```

You should get 1.475261. How would you interpret that value?

Let's create a smaller data frame (called `college_ratios`) that selects only the variables needed to calculate this ratio. We'll look at the colleges with the ten highest ratios and create a new variable called `rank` that captures each school's rank (where the institution with the highest ratio is ranked #1):

```{r summarizing ratios}
college_ratios <- colleges %>% 
        arrange(desc(ratio)) %>%
        mutate(rank = row_number()) %>%
        top_n(-10) %>%
        select(rank, name, ratio, par_top1pc, par_bottom60pc)

college_ratios
```

So Middlebury's ratio of 1.4571 is the fourth highest in this sample. That ratio can be interpreted as: "Middlebury has about 1.5 students from the top 1% of the income distribution for every student from the bottom 60%."

## Mobility Measures

Now let's turn to the two measures that are the focus of the full Chetty paper. The first variable is a proxy for how accessible colleges are to students from the bottom 20% of the income distribution; the variable is called the `access_rate` and it measures the proportion of students from the bottom 20% of the income distribution. The second variable measures the proportion of these students from the bottom 20% of the income distribution who make it to the top 20% of the students' income distribution by the time they are approximately 30 years old; this variable is called the `success_rate`. 

Let's find these two measures for Middlebury (note that we are using the `colleges` data frame again so we have all the variables):

```{r access rate and success rate}
colleges$access_rate[colleges$name=="Middlebury College"]
colleges$success_rate[colleges$name=="Middlebury College"]
```

What do these values mean? For the pooled years in this sample, 2.3% of Middlebury students came from the bottom quintile of the parents' income distribution, and 54.6% *of those students* ended up in the top quintile of the students' income distrubtion in early adulthood.

The product of these values is the proportion of *all* Middlebury students who are from the lowest income quintile and who make it to the top income quintile. Chetty et al call this measure the `mobility rate`. Let's calculate this measure for all colleges and find the value for Middlebury:

```{r calculating the mobility rate}
colleges$mobility_rate <- colleges$access_rate * colleges$success_rate

colleges$mobility_rate[colleges$name=="Middlebury College"]
```

So in this sample, 1.3% of Middlebury students grow up in the bottom 20% of the income distribution but end up in the top 20%.

Where does Middlebury's value fall in the overall distribution of mobility rates at four-year colleges?

```{r summarizing the mobility rate}
summary(colleges$mobility_rate)
```

Looks like Middlebury's mobility rate falls just below the median, which tells us that more than 50% of the other institutions in this sample have higher mobility rates than Middlebury.

## Correlates of Success Rate

Chetty et al use several other college characteristics - gathered from the Common Data Set, the College Scorecard, and other IPEDS databases - to see how they are related to the success, access, and mobility rates. Let's look at cost of attendance first. Would you expect colleges with higher "sticker prices" to have higher or lower success rates on average? 

Let's find the correlation using the `cor()` function, which will tell us how the distributions of these values tend to move together. The `use = "complete.obs"` option asserts that we want to drop observations missing data on one or both of the variables.

```{r correlation}
cor(colleges$sticker_price_2013, colleges$success_rate, 
    use = "complete.obs")
```

Correlations can be negative (between 0 and -1) or positive (between 0 and 1). The correlation we calcuated is positive. In words, this correlation means that colleges with higher sticker prices tend to have higher success rates, on average. A negative correlation would have told us that colleges with higher sticker prices tend to have lower success rates, on average.

Correlations can be described by their sign (positive or negative) and their strength. There are no strict rules for measuring strength, but here is one summary you can use for this workshop:

- Positive correlations between 0 and .19 or negative correlations between 0 and -.19 = no association

- Positive correlations between .20 and .29 or negative correlations between -.20 and -.29 = weak association

- Positive correlations between .30 and .49 or negative correlations between -.30 and -.49 = moderate association

- Positive correlations between .50 and .69 or negative correlations between -.50 and -.69 = strong association

- Positive correlations greater than or equal to .70 or negative correlations less than or equal to -.70 = very strong association


A complete description of the correlation between sticker price and success rate would say that these two variables have a moderately strong positive association.

## Visualizing Associations

The relationship between these two variables might be easier to interpret in a scatterplot with a "line of best fit." After setting our x and y variables, R knows to make a scatterplot because we ask for a `geom_point()` and knows to add a line because we add `geom_smooth(method = lm)`:

```{r scatterplot , message = FALSE, warning = FALSE}
plot_sticker_price <- ggplot(colleges, aes(x = sticker_price_2013, 
                                           y = success_rate))

plot_sticker_price + geom_point() + geom_smooth(method = lm)
```

How do you think the relationship between sticker price and success rate could differ between public and private colleges?

Here are the correlations for public colleges (with a value of 1 for the public variable) vs private colleges (with a value of 0 for the public variable):

```{r public private correlations}
colleges %>% group_by(public) %>% 
          summarize(correlation = cor(sticker_price_2013, success_rate,
                                      use = "complete.obs")) %>% 
     select(public, correlation)
```

And here is a scatterplot with private colleges in pink and public colleges in blue:

```{r public private scatterplot, message = FALSE, warning = FALSE}
plot_sticker_price <- ggplot(colleges, aes(x = sticker_price_2013, 
                                           y = success_rate, 
                                           color = public))

plot_sticker_price + geom_point() + geom_smooth(method = lm)
```

Looks like private colleges have the highest sticker prices, as well as the highest *and lowest* success rates. 

Where is Middlebury's point in this distribution? There are a few ways to do this. The easiest is to use the `gghighlight` package which needs to be loaded:

```{r load gghighlight}
library(gghighlight)
```

Now we add a `gghighlight` command to the plot saying we want to highlight the point for Middlebury and add a label (with the label specifying the value of the `name` variable).

```{r gghighlight plot, warning = FALSE}
plot_sticker_price <- ggplot(colleges, aes(x = sticker_price_2013, 
                                           y = success_rate))
plot_sticker_price + geom_point() + 
     gghighlight(name=="Middlebury College", label_key = name) +
     geom_smooth(data = colleges, method = lm, color = "red")
```


\pagebreak

## Workshop Report

Answer the following questions in a separate document (not in R). Upload a PDF of your document to Canvas by 8:00 PM (Middlebury Time) on Friday, May 1, 2020. You are free to collaborate with classmates on this workshop report, but each student must upload their own submission.

1. Choose two of the following variables and find how each of them is correlated with the **success rate**. Select two measures you expect might have different correlations with the success rate (one positive and one negative, or one strongly positive and one weakly positive, for example). Interpret both correlations in words. For one of the variables, include a scatterplot showing how it is related to the success rate.

The available variables are:

`admit_rate_2013`: Proportion of undergraduate applicants accepted for the class that entered in Fall 2013;

`asian_or_pacific_share_fall_2000`: Share of Asian or Pacific Islander undergraduate students in Fall 2000;

`average_faculty_salary_2013`: Average salary of instructional faculty;

`black_share_fall_2000`: Share of Black undergraduate students in Fall 2000;

`endowment_pc_2000`: Endowment assets per student in 2000;

`female`: Proportion of students who self-identify as female;

`graduation_rate_2013`: Proportion of students graduating within 6 years of entry;

`hisp_share_fall_2000`: Share of Hispanic undergraduate students in Fall 2000;

`instructional_expenditures_pc_2012`: Instructional expenditures per student in 2000;

`ipeds_enrollment_2013`: Total IPEDS undergraduate enrollment (full time and part time) in 
Fall 2013;

`pct_arthuman_2000`: Arts and Humanities major share in 2000;

`pct_socialscience_2000`: Social Sciences major share in 2000;

`pct_stem_2000`: STEM major share in 2000;

`scorecard_netprice_2013`: Net cost of attendance for students from the bottom 20% of the income distribution in 2013;

`sat_avg_2013`: Average SAT scores (scaled to 1600) in 2013; values for this variable are missing for institutions that do not require the SAT


2. What is another variable that is not included in this dataset that you expect would have a *positive relationship* with the **access rate**? What could explain this relationship?

3. What is another variable that is not included in this dataset that you expect would have a *negative relationship* with the **access rate**? What could explain this relationship?

4. Think back to the **mobility rate** and how it is calculated. In a few sentences, discuss some advantages and disadvantages in using this single measure to compare colleges.